---
title: "Preprocessing"
author: "Matthew Angel"
date: "12/13/2021"
output:
  rmarkdown::html_document:
    toc: true
    code_folding: hide

vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---
<style type="text/css">

body{ /* Normal  */
      font-size: 12px;
  }
td {  /* Table  */
  font-size: 8px;
}
h1.title {
  font-size: 38px;
  color: DarkRed;
}
h1 { /* Header 1 */
  font-size: 28px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Data preprocessing: F60

```{r preprocessing}

  suppressMessages(library("qdap"))
  suppressMessages(library("stringr"))
  suppressMessages(library("tidyverse"))

  df <- read.csv("input/f60_cytokine_concentrations.csv", check.names = FALSE)

  # Flip dataset
  row.names(df) <- df[["Vial_Label"]]
  df[["Vial_Label"]] <- NULL
  
  converter_in  <- c("α","β","/")
  converter_out <- c("alpha","beta","_")
  
  colnames(df) <- multigsub(converter_in,converter_out,colnames(df))
  
  df <- t(as.matrix(df))
  class(df) <- "numeric"
  
  df <- log2(df + 1)
  df <- as.data.frame(df) %>% rownames_to_column("Gene")
  write.table(df, file = "output/f60_formatted_data.csv", sep = ",", row.names = FALSE, quote = FALSE)
  
  cat("Genes in dataset:")
  df$Gene
  
  cat("Samples in dataset:")
  colnames(df)[-1]
  
```

# Diff Counts

```{r diff_counts}

    annot <- read.csv("input/f60_metadata.csv")

    colnames(df) <- gsub("_G\\d_","_grouped_",colnames(df))
    annot$sample_id <- gsub("_G\\d_","_grouped_",annot$sample_id)
    
    annot <- annot %>% filter(sample_id %in% colnames(df))
    
    contrasts <- c("d1-d0","d3-d0","d7-d0")
    annot$group <- 'grouped'

    annot$vaccination <- factor(annot$vaccination)
    annot$group <- factor(as.character(annot$group))
    
    for(vac in levels(annot$vaccination)){
        annot.v <- annot %>% filter(vaccination == vac)
        for(g in levels(annot$group)){
            annot.g <- annot.v %>% filter(group == g)
            for(n in seq_along(contrasts)){
                c <- contrasts[n]
                test <- unlist(str_split(c,"-"))[1]
                ref  <- unlist(str_split(c,"-"))[2]

                annot.c <- annot.g %>% filter(timepoint %in% c(test,ref))
                samples_in_contrast <- annot.c$sample_id

                df.c <- df[ , c("Gene",samples_in_contrast)]
                row.names(df.c) <- df$Gene
                df.c$Gene <- NULL

                test.samples <- annot.c %>% filter(timepoint %in% test) %>% pull(sample_id)
                ref.samples  <- annot.c %>% filter(timepoint %in% ref)  %>% pull(sample_id)
 
                df.t <- as.data.frame(t(df.c))

                df.t$animal_id   <- apply(array(row.names(df.t)), 1, function(z) unlist(str_split(z,"_"))[1])

                df.ret <- as.data.frame(matrix(nrow=length(row.names(df.c)),ncol=length(unique(df.t$animal_id)),dimnames=list(row.names(df.c),paste0(unique(df.t$animal_id),"_",vac,"_",g,"_",test,"-",ref))))
                for(i in 1:nrow(df.ret)){
                    gene <- row.names(df.ret)[i]
                    for(j in 1:ncol(df.ret)){
                        col.nam <- colnames(df.ret)[j]
                        animal_id <- unlist(str_split(col.nam,"_"))[1]
                        contrast <- unlist(str_split(col.nam,"_"))[4]
                        t.level <- unlist(str_split(contrast,"-"))[1]
                        r.level <- unlist(str_split(contrast,"-"))[2]

                        test.sample <- paste(animal_id,g,vac,t.level,sep="_")
                        ref.sample  <- paste(animal_id,g,vac,r.level,sep="_")

                        #Do we actually have both levels?
                        if(!all(c(test.sample,ref.sample) %in% row.names(df.t))){
                            df.ret[i,j] <- NA
                            next
                        }

                        #Are either NA?
                        if( is.na(df.t[test.sample,gene]) | is.na(df.t[ref.sample,gene]) ){
                                df.ret[i,j] <- NA
                            next                
                        }

                        df.ret[i,j] <- df.t[test.sample,gene] - df.t[ref.sample,gene]
                    }
                }

                df.ret <- df.ret %>% rownames_to_column("Gene")
                if(n == 1 & vac == "V1" & g == "grouped"){
                    df.all <- df.ret
                }else{
                    df.all <- merge(df.all,df.ret,by="Gene")
                }
            }
        }
    }

  write.table(df.all, file = "output/f60_diff_counts.csv", sep = ",", row.names = FALSE, quote = FALSE)
```

# Diff Metadata

```{r diff_metadata}


    sample_ids <- colnames(df.all)
    sample_ids <- sample_ids[ sample_ids != "Gene"]

    animal_ids <- apply(array(sample_ids), 1, function(z) unlist(str_split(z, "_"))[1])
    vaccination <- apply(array(sample_ids), 1, function(z) unlist(str_split(z, "_"))[2])
    grp <- apply(array(sample_ids), 1, function(z) unlist(str_split(z, "_"))[3])
    contrast <- apply(array(sample_ids), 1, function(z) unlist(str_split(z, "_"))[4])

    df <- data.frame(sample_id = sample_ids, animal_id = animal_ids, group = grp, vaccination = vaccination, contrast = contrast)

    write.table(df, file = "output/f60_diff_metadata.csv", sep = ",", row.names = FALSE, quote = FALSE)
    
    df
```

# Session Info

```{r session_info}

sessionInfo()

```
